The algebraic hierarchy in type theory using type classes

### 2010-06-22 jar.tex
small changes

To this end, we have formalized some of the theory of multisorted universal algebra and equational theories. We chose not to revive existing old (and somewhat stale) formalizations ~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because an important aim for us has been to find out what level of elegance, convenience, and integration can be achieved by leveraging the state of the art in Coq facilities (of which type classes are the most important example).

@@

  Class AlgebraOps (σ: Signature) (carriers: sorts σ → Type)
    := algebra_op: Π o: operation σ, op_type carriers (σ o).
\end{lstlisting}
Because our carriers will normally be equipped with a setoid equality, we further define the predicate class \lstinline|Algebra|, stating that each of the operations respects the setoid equality on the carriers:
\begin{lstlisting}
  Class Algebra `{Π a, Equiv (carriers a)} `{AlgebraOps σ carriers}: Prop :=
      { algebra_setoids:> Π a, Setoid (carriers a)
      ; algebra_propers:> Π o: σ, Proper (=) (algebra_op o) }.
\end{lstlisting}
The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an |Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.

@@

Algebras are insufficient to characterize structures such as semirings and rings. For that, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
\begin{lstlisting}
Record EquationalTheory :=
  { et_sig:> Signature
  ; et_laws:> EqEntailment et_sig → Prop }.
\end{lstlisting}
An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar. A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper,

@@

What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the ``variety''). Since we need a type for the objects in the |Category|, at this point we have no choice but to bundle components and proof together in a record:
\begin{lstlisting}
  Variable et: EquationalTheory.

@@

For the main theorem, we now bundle the quotient algebra and the subalgebra into records akin to \lstinline|ObjectInVariety| from section~\ref{varieties}:
\begin{lstlisting}
  Definition quot_obj: algebra.Object σ := algebra.object σ A (algebra_equiv:=(~)).
  Definition subobject: algebra.Object σ := algebra.object σ (ua_subalgebraT.carrier image).
\end{lstlisting}
Here, \lstinline|algebra| is the module defining the bundled algebra record \lstinline|Object| with constructor \lstinline|object|. \lstinline|ua_subalgebraT| is the module for constructing subalgebras.

@@

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Class Naturals A {e plus mult zero one} `{!NaturalsToSemiRing A} :=
    { naturals_ring:> @SemiRing A e plus mult zero one
    ; naturals_to_semiring_mor:> Π `{SemiRing B}, SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound (think unary, binary, factor multisets,
arrays of native machine words, etc.).

>>>>

To this end, we have formalized some of the theory of multisorted universal algebra and equational theories. We chose not to revive existing old (and somewhat stale) formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because an important aim for us has been to find out what level of elegance, convenience, and integration can be achieved by leveraging the state of the art in Coq facilities (of which type classes are the most important example).

@@

  Class AlgebraOps (σ: Signature) (carriers: sorts σ → Type)
    := algebra_op: Π o: operation σ, op_type carriers (σ o).
\end{lstlisting}
Because our carriers will normally be equipped with a setoid equality, we further define the predicate class \lstinline|Algebra|, stating that each of the operations respects the setoid equality on the carriers:
\begin{lstlisting}
  Class Algebra `{Π a, Equiv (carriers a)} `{AlgebraOps σ carriers}: Prop :=
      { algebra_setoids:> Π a, Setoid (carriers a)
      ; algebra_propers:> Π o: σ, Proper (=) (algebra_op o) }.
\end{lstlisting}
The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.

@@

To characterize structures such as semirings and rings, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
\begin{lstlisting}
Record EquationalTheory :=
  { et_sig:> Signature
  ; et_laws:> EqEntailment et_sig → Prop }.
\end{lstlisting}
An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar. A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper,

@@

What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the ``variety''). Since we need a type for the objects in the \lstinline|Category|, at this point we have no choice but to bundle components and proof together in a record:
\begin{lstlisting}
  Variable et: EquationalTheory.

@@

For the main theorem, we now bundle the quotient algebra and the subalgebra into records akin to \lstinline|ObjectInVariety| from section~\ref{varieties}:
\begin{lstlisting}
  Definition quot_obj: algebra.Object σ :=
    algebra.object σ A (algebra_equiv:=(~)).
  Definition subobject: algebra.Object σ :=
    algebra.object σ (ua_subalgebraT.carrier image).
\end{lstlisting}
Here, \lstinline|algebra| is the module defining the bundled algebra record \lstinline|Object| with constructor \lstinline|object|. The module \lstinline|ua_subalgebraT| constructs subalgebras.

@@

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Class Naturals A {e plus mult zero one} `{!NaturalsToSemiRing A} :=
    { naturals_ring:> @SemiRing A e plus mult zero one
    ; naturals_to_semiring_mor:> Π `{SemiRing B}, SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse and non-contrived representations of the natural numbers are abound ( unary, binary, factor multisets, arrays of native machine words, etc.).
